---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dev = "ragg_png"
)
```

# chronogram

<!-- badges: start -->

[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable) [![R-CMD-check](https://github.com/FrancisCrickInstitute/chronogram/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/FrancisCrickInstitute/chronogram/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

The goal of `chronogram` is to "cast" and annotate metadata, laboratory and clinical data into a tidy-like data structure. This bridges between a LIMS / database style data warehouse and data that is ready for interrogation to test biological hypotheses.

`chronogram` was designed during the SARS-CoV-2 pandemic (2019-). However, it is pathogen, vaccine and symptoms agnostic.

------------------------------------------------------------------------

## Installation

Install the current version from [GitHub](https://github.com/):

``` r
# install.packages("devtools")
devtools::install_github("FrancisCrickInstitute/chronogram")
```

If you have not installed packages from github before, you will to [setup your GitHub account to interact with R](https://usethis.r-lib.org/articles/git-credentials.html#practical-instructions).

------------------------------------------------------------------------

## Why should I use `chronogram`?

-   To aggregate study data **regularly 🕓**, and **repetitively 🔁**. Perhaps your study has rolling recruitment, ongoing data generation or incremental analysis. Outsource that effort to `chronogram`.

-   To **reproducibly aggregate** data within and **across several studies and users 👩‍💻👨‍💻**. Stop troubleshooting joins by hand.

-   To provide a **versatile** data shape **poised** for **new or follow-up analyses** without needing re-aggregation 🛫.

***When shouldn't I use `chronogram`?***

Your study is **completed**. You have assembled a clean, de-duplicated and fully annotated data object. You have **finished all data analysis**. Congratulations! 🥳 Don't reinvent the wheel here.

------------------------------------------------------------------------

## How do I use `chronogram`?

The `chronogram` workflow can be divided into assembly, annotation and finally, filtering, windowing and selecting data for a specific analysis.

### chronogram assembly

-   `cg_assemble()` combines cleaned metadata, experimental data, and a range of calendar dates into a chronogram.

-   `cg_add_experiment()` allows the adding of further experiments

Further details:

-   [assembly](articles/assembly.html) vignette for a step-by-step guide, or the [quickstart](articles/chronogram.html).

-   [SQL vignette](articles/SQL_assembly.html) explains `chronogram` assembly from an SQL database.

-   An introduction to the [chronogram class](articles/chronogram_class.html)

### chronogram annotation

#### Annotate vaccines

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=3}
library(chronogram)
library(dplyr)
library(ggplot2)

data(built_smallstudy)
cg <- built_smallstudy$chronogram
cg <- cg_add_experiment(cg, built_smallstudy$infections_to_add)

cg <- cg_annotate_vaccines_count(
  cg,
  ## the prefix to the dose columns: ##
  dose = dose,
  ## the output column name: ##
  dose_counter = dose_number,
  ## the prefix to the date columns: ##
  vaccine_date_stem = date_dose,
  ## use 14d to 'star' after a dose ##
  intermediate_days = 14
) %>%
  mutate(dose_number = factor(dose_number,
    levels = c(
      "0",
      "1star",
      "1",
      "2star",
      "2"
    )
  ))

## plot over time ##
cg %>%
  ggplot(
    aes(
      x = calendar_date,
      y = elig_study_id,
      fill = dose_number
    )
  ) +
  geom_tile(height = 0.5) +
  scale_fill_grey(end = 0.2, start = 0.8) +
  theme_bw()

```

Label each day for each participant with the number of doses they have received, including support for a lag period between the reciept of a dose and its immunological priming effect. [Annotate vaccines here](articles/annotate_vaccines.html).

#### Annotate infection episodes

```{r echo=FALSE, warning=FALSE, message=FALSE, results="none"}
library(chronogram)
library(dplyr)
library(ggplot2)

data(built_smallstudy)
cg <- built_smallstudy$chronogram
cg <- cg_add_experiment(cg, built_smallstudy$infections_to_add)

cg <- cg_annotate_episodes_find(
  cg,
  infection_cols = c("LFT", "PCR", "symptoms"),
  infection_present = c("pos", "Post", "^severe"),
  episode_days = 90
)

cg <- cg %>%
  mutate(
    episode_variant =
      case_when(
        # "is an episode" & "PCR positive" -> Delta #
        (!is.na(episode_number)) & PCR == "Pos" ~ "Delta",
        # "is an episode" & "PCR unavailable" -> Anc/Delta #
        (!is.na(episode_number)) & PCR == "not tested" ~ "Anc/Alpha"
      )
  )

cg <- cg %>%
  cg_annotate_episodes_fill(col_to_fill = episode_variant,
                            col_to_return = episode_variant_filled)

## plot over time ##
cg %>%
  ggplot(
    aes(
      x = calendar_date,
      y = elig_study_id,
      fill = episode_variant_filled
    )
  ) +
  geom_tile(height = 0.1, width = 5) +
  geom_point(data = . %>% 
               ## filter to LFT results that are present ##
               filter(!is.na(LFT)) %>% 
               ## preceed with "LFT" to allow a single colour guide ##
               mutate(LFT = paste("LFT", LFT)),
             aes(col = LFT),
             shape = "I",
             size = 4,
             position = position_nudge(y=0.4)) +
  ## repeat LFT approach for PCR
  geom_point(data = . %>% 
               filter(!is.na(PCR)) %>%
               mutate(PCR = paste("PCR", PCR)),
             aes(col = PCR),
             shape = "I",
             size = 4,
             position = position_nudge(y=0.6)) +
  ## repeat LFT approach for symptoms
  geom_point(data = . %>% 
               filter(!is.na(symptoms)) %>%
               mutate(symptoms = paste("symptoms", symptoms)),
             aes(col = symptoms),
             shape = "I",
             size = 4,
             position = position_nudge(y=0.2)) +
  ## swap the fill scale, and stop colour being included in this guide ##
  scale_fill_grey(name = "infection episode",
                  na.translate = FALSE,
                  na.value = NA, 
                  
                  guide = guide_legend(override.aes = list(color = NA))) +
  ## swap the colour scale ##
  scale_color_brewer(type = "qual",palette = 2,
                     name = "nasopharyngeal testing & symptoms") + 
  theme_bw() + 
  theme(legend.position = "right",
        legend.direction = "vertical")

```

Symptoms, point-of-care tests, and laboratory tests of infection rarely occur on exactly the same study day. `chronogram` finds, fills and annotates these tests and symptoms into episodes of infection. [Annotate episodes here](articles/annotate_episodes.html).

#### Annnotate exposures

```{r echo=FALSE, warning=FALSE, message=FALSE,results="none", fig.height=7}
library(chronogram)
library(dplyr)
library(ggplot2)
library(patchwork)

cg <- cg_annotate_vaccines_count(
  cg,
  ## the prefix to the dose columns: ##
  dose = dose,
  ## the output column name: ##
  dose_counter = dose_number,
  ## the prefix to the date columns: ##
  vaccine_date_stem = date_dose,
  ## use 14d to 'star' after a dose ##
  intermediate_days = 14
) %>%
  mutate(dose_number = factor(dose_number,
    levels = c(
      "0",
      "1star",
      "1",
      "2star",
      "2"
    )
  ))

cg_exposures <- cg %>% cg_annotate_exposures_count(
  episode_number = episode_number,
  dose_number = dose_number,
  ## we have not considered episodes of seroconversion
  N_seroconversion_episode_number = NULL
)


cg_exposures <- cg_exposures %>%
  mutate(
    episode_variant_summarised =
      episode_variant_filled
  ) %>%
  cg_annotate_antigenic_history(
    episode_number = episode_number,
    dose_number = dose_number,
    episode_variant_summarised = episode_variant_summarised,
    ag_col = antigenic_history
  )

## Plot ##
top_panel <- cg_exposures %>%
  select(calendar_date, 
         exposure_number,
         elig_study_id,
         antigenic_history) %>%
  ggplot(aes(
    x = calendar_date, y = exposure_number,
    col = elig_study_id
  )) +
  geom_line() +
  facet_grid(antigenic_history ~ .)


swimmers_panel <- cg_plot_meta(cg_exposures,
  visit = serum_Ab_S
) +
  ## set the axes to match top_panel ##
  xlim(
    min(cg_exposures$calendar_date),
    max(cg_exposures$calendar_date)
  ) +
  scale_y_discrete(limits = factor(c(3, 2, 1)))

top_panel / swimmers_panel & theme_bw() &
  theme(
    legend.position = "bottom",
    strip.text.y = element_text(angle = 0),
    strip.background = element_blank(),
    panel.grid.minor = element_blank()
  )
```

After annotating vaccines and infection episodes, these can be combined to [annotate exposures](articles/annotate_exposures.html) - encounters with antigen from either infection or vaccination.

------------------------------------------------------------------------

### chronogram filtering, window and select

-   `dplyr::filter()` to filter a chronogram based on metadata (eg vaccine formulation)

-   `cg_window_by_metadata()` to window around an event such as 14 days after each participant's vaccine

-   `cg_window_by_episode()` picks a window around infection episodes

See these functions at work in our [brief primer](articles/stats.html) demonstrating how to a pass `chronogram` to a variety of statistical tests.

------------------------------------------------------------------------
