---
title: "Annotate exposures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Annotate exposures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{stringr}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 8
)
```


## Introduction

The chronogram package provides a family of functions to annotate a chronogram. These all start `cg_annotate_`. This vignette explains how to use these annotation functions. Before using this vignette, consult the `vignette("assembly")`.

This vignette demonstrates annotation of exposures. Exposures can be either from vaccines, or infection episodes - these require annotation first. See: `vignette("annotate vaccines")` & `vignette("annotate episodes")`.

## Setup

```{r setup}
library(chronogram)
library(dplyr)
library(ggplot2)
library(patchwork)
```

We will use the example pre-built chronogram, introduced in the `vignette("assembly")`, with vaccines and episodes annotated: `vignette("annotate vaccines")` & `vignette("annotate episodes")`. The code chunk below is discussed in those vignettes.

```{r}
data(built_smallstudy)
cg <- built_smallstudy$chronogram
infections_to_add <-  built_smallstudy$infections_to_add

## add to chronogram
cg <- cg_add_experiment(
  cg,
  infections_to_add
)

## annotate vaccines
cg <- cg_annotate_vaccines_count(
  cg,
  ## the prefix to the dose columns: ##
  dose = dose,
  ## the output column name: ##
  dose_counter = dose_number,
  ## the prefix to the date columns: ##
  vaccine_date_stem = date_dose,
  ## use 14d to 'star' after a dose ##
  intermediate_days = 14
)

## annotate episodes
cg <- cg %>%
  cg_annotate_episodes_find(
  infection_cols = c("LFT", "PCR", "symptoms"),
  infection_present = c("pos", "Post", "^severe")
) %>%
  mutate(
    episode_variant =
      case_when(
        # "is an episode" & "PCR positive" -> Delta #
        (!is.na(episode_number)) & PCR == "Pos" ~ "Delta",
        # "is an episode" & "PCR unavailable" -> Anc/Delta #
        (!is.na(episode_number)) & PCR == "not tested" ~ "Anc/Alpha"
      )
  ) %>%
  cg_annotate_episodes_fill(
    col_to_fill = episode_variant,
    col_to_return = episode_variant_filled,
    .direction = "updown"
  )
```

## Outline

Annotation is required to allow the selection sub-cohorts of individuals (and corresponding dates) that are relevant to test your biological hypothesis.

Exposures can be either from vaccine doses or from infection episodes. `cg_annotate_exposure_count()` provides a cumulative counter for each individual's personal history. `cg_annotate_antigenic_history()` returns a text string summarising the sequence of encounters.


### Worked example

`cg_annotate_exposure_count()` takes the column names for episode number, dose number and seroconversion episode numbers, to calculate a per day, per individual running count of exposures.

`cg_annotate_antigenic_history()` provides a character vector of length 1 to summarise each person's course over the study.

These are both best explored with an example, and a plot.


```{r}
cg_exposures <- cg %>% cg_annotate_exposures_count(
  episode_number = episode_number,
  dose_number = dose_number,
  ## we have not considered episodes of seroconversion
  N_seroconversion_episode_number = NULL
)


cg_exposures <- cg_exposures %>%
  mutate(
    episode_variant_summarised =
      episode_variant_filled
  ) %>%
  cg_annotate_antigenic_history(
    episode_number = episode_number,
    dose_number = dose_number,
    episode_variant_summarised = episode_variant_summarised,
    ag_col = antigenic_history
  )

## Plot ##
top_panel <- cg_exposures %>%
  select(calendar_date, 
         exposure_number,
         elig_study_id,
         antigenic_history) %>%
  ggplot(aes(
    x = calendar_date, y = exposure_number,
    col = elig_study_id
  )) +
  geom_line() +
  facet_grid(antigenic_history ~ .)


swimmers_panel <- cg_plot_meta(cg_exposures,
  visit = serum_Ab_S
) +
  ## set the axes to match top_panel ##
  xlim(
    min(cg_exposures$calendar_date),
    max(cg_exposures$calendar_date)
  ) +
  scale_y_discrete(limits = factor(c(3, 2, 1)))

top_panel / swimmers_panel & theme_bw() &
  theme(
    legend.position = "bottom",
    strip.text.y = element_text(angle = 0),
    strip.background = element_blank(),
    panel.grid.minor = element_blank()
  )
```

The plot above shows how each infection, or vaccination impacts the exposure number. For participant 3, their only exposures are vaccination, whereas both participants 1 and 2 have additional exposures from infection. By the end of this example, participant 2 has experienced 4 encounters with Spike, participant 1 has had 3 encounters and participant 3 just two from vaccination alone.

# Summary

This vignette has provided examples of the cg_annotate family in action. If you are conducting a multi-pathogen study (RSV, flu, covid), then run a set of cg_annotate family functions for each pathogen - and you may wish to prefix the output columns eg `RSV_`,  `flu_` & `covid_`. As these have differing considerations for eg variants, chronogram leaves the cg_annotate family without an overall wrapper to let users easily omit unneeded annotations.

## SessionInfo

```{r}
sessionInfo()
```
