---
title: "Annotate vaccines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Annotate vaccines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{stringr}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 8
)
```


## Introduction

The chronogram package provides a family of functions to annotate a chronogram. These all start `cg_annotate_`. This vignette explains how to use these annotation functions. Before using this vignette, consult the `vignette("assembly")`.

## Setup

```{r setup}
library(chronogram)
library(dplyr)
library(ggplot2)
library(patchwork)
```

We will use the example pre-built chronogram, introduced in the `vignette("assembly")`, and add on some example infection data.

```{r}
data(built_smallstudy)
cg <- built_smallstudy$chronogram
infections_to_add <-  built_smallstudy$infections_to_add

## add to chronogram
cg <- cg_add_experiment(
  cg,
  infections_to_add
)
```

## Vaccine annotation

Annotation is required to allow the selection sub-cohorts of individuals (and corresponding dates) that are relevant to test your biological hypothesis.

For vaccines, use `cg_annotate_vaccines_count()`

-- adds a column which counts the number of vaccines each participant has received over time.

-- includes a "star" system, to allow the first few days after a vaccine to be easily identified. For example, 24hrs after the first dose of a vaccine ("1star") is probably not biologically reflect of that dose's effect. The user can set the value of days to "star" to suit their analysis.

`cg_annotate_vaccines_count()` requires that metadata columns follow this pattern:

- dose_1, dose_2, dose_3, ..., dose_i

- date_dose_1, date_dose_2, date_dose_3, ..., date_dose_i

The trailing digit reflects the dose number for both `dose_i` and `date_dose_i`.

The {dose} and {date_dose} prefixes should be provided to the function, as shown in the chunk below. You might envisage a study with {sarscov2_dose}, {sarscov2_date_dose}, {influenza_dose} & {influenza_date_dose} for which two runs of `cg_annotate_vaccines_count()` would be needed.

### Worked example

```{r}
cg <- cg_annotate_vaccines_count(
  cg,
  ## the prefix to the dose columns: ##
  dose = dose,
  ## the output column name: ##
  dose_counter = dose_number,
  ## the prefix to the date columns: ##
  vaccine_date_stem = date_dose,
  ## use 14d to 'star' after a dose ##
  intermediate_days = 14
)

## plot over time ##
cg %>%
  ggplot(
    aes(
      x = calendar_date,
      y = elig_study_id,
      fill = dose_number
    )
  ) +
  geom_tile(height = 0.5) +
  scale_fill_grey(end = 0.2, start = 0.8) +
  theme_bw()
```

In the above plot, the character vector `dose_number` is coerced to a factor. The resulting levels of the factor are counter-intuitive. To improve the plot, you can manually specify the factor: 

```{r}
cg %>%
  mutate(dose_number = factor(dose_number,
    levels = c(
      "0",
      "1star",
      "1",
      "2star",
      "2"
    )
  )) %>%
  ggplot(
    aes(
      x = calendar_date,
      y = elig_study_id,
      fill = dose_number
    )
  ) +
  geom_tile(height = 0.5) +
  scale_fill_grey(end = 0.2, start = 0.8) +
  theme_bw()
```

The `dose_number` column is intentionally returned to `cg` as character vector rather than converting to a factor. There is the possibility for mishandling if a factor: `as.numeric(dose_number) == 1`, refers to the situation `dose == 0`, and `as.numeric(dose_number) == 2`, refers to the situation `dose == 1`.

# Summary

This vignette has provided examples of the cg_annotate family in action. If you are conducting a multi-pathogen study (RSV, flu, covid), then run a set of `cg_annotate` family functions for each pathogen - and you may wish to prefix the output columns eg `RSV_`,  `flu_` & `covid_`.

## SessionInfo

```{r}
sessionInfo()
```
