---
title: "Quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```

This vignette is for the impatient and showcases the features of the `chronogram` package.

## Load libraries

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(chronogram)
```

## Construct a chronogram

Use `cg_assemble()` to create a chronogram from cleaned input data.

```{r}
## Fictional example data ##
data(smallstudy)


cg <- cg_assemble(
  start_date = "01012020",
  end_date = "10102021",
  ## the provided metadata ##
  metadata = smallstudy$small_study_metadata,
  ## the column name in the metadata that contains participant IDs ##
  metadata_ids_col = elig_study_id,
  ## column name for dates ##
  calendar_date_col = calendar_date,
  ## the provided experiment data (we have 1 assay, so a list of 1) ##
  experiment_data_list = list(smallstudy$small_study_Ab)
)

```

`print()`, `glimpse()` and `summary()` help explore the resulting chronogram. (`View()` works too, but not shown in vignette).

```{r}
print(cg)

glimpse(cg)

summary(cg)

```

## Plot trajectories

Plot antibody responses over time for each individual using `cg_plot()`.

```{r}
traj <- cg_plot(cg,
  y_values = serum_Ab_S
)

traj
```

Plot a swimmers plot to visualise an individual's sequence of events within the study, using `cg_plot_meta()`.

```{r}
swimmer <- cg_plot_meta(cg,
  visit = serum_Ab_S
)

swimmer
```

### Customise plots

Using `ggplot2` or `patchwork` customisations.

```{r}
lay <- "
a
a
a
b
"
(
  (traj + labs(y = "anti-S IgG") + scale_y_log10()) /
    (swimmer + theme(axis.text.y = element_blank()))
) +
  plot_layout(design = lay) &
  plot_annotation(tag_levels = "A") &
  theme_bw(base_size = 8) &
  theme(legend.position = "none") &
  theme(plot.tag = element_text(size = 16, face = "bold")) &
  labs(x = "date")
```

## Annotate a chronogram

There are a suite of functions to find, label and count infection episodes. These aggregate across different experiment data columns, such as symptoms, PCRs, viral sequencing. See the [annotation vignette](docs/articles/annotation.html).

## Window a chronogram

Here, we pick dates in relation to dose 2, using `cg_window_by_metadata()`.

```{r}
## upto 50d after dose 2 ##
cg_after_dose_2 <- cg_window_by_metadata(cg,
  windowing_date_col = date_dose_2,
  preceding_days = 0,
  following_days = 50
)

## upto 21d before dose 2 ##
cg_before_dose_2 <- cg_window_by_metadata(cg,
  windowing_date_col = date_dose_2,
  preceding_days = 21,
  following_days = 0
)


cg_after_dose_2 <- cg_after_dose_2 %>% mutate(cohort = "post D2")

cg_before_dose_2 <- cg_before_dose_2 %>% mutate(cohort = "pre D2")

cg_dose_2 <- bind_rows(cg_after_dose_2, cg_before_dose_2)

cg_dose_2 <- cg_dose_2 %>% mutate(
  cohort =
    factor(cohort,
      levels = c(
        "pre D2",
        "post D2"
      )
    )
)
```

### Re-plot the windowed data

```{r}
traj_2 <- cg_dose_2 %>%
  cg_plot(.,
    y_values = serum_Ab_S
  )

pd <- position_dodge(0.4)

violin_2 <- cg_dose_2 %>%
  ## geom_line fails if lots of NA rows provided ##
  filter(!is.na(serum_Ab_S)) %>%
  ggplot(aes(
    x = cohort, y = serum_Ab_S,
    group = elig_study_id
  )) +
  geom_point(position = pd) +
  geom_line(position = pd)

swimmer_2 <- cg_plot_meta(cg_dose_2,
  visit = serum_Ab_S
)


lay2 <- "
aac
aac
aa#
bb#
"

(
  (traj_2 +
    labs(y = "anti-S IgG", x = "date") +
    scale_y_log10()) +
    (swimmer_2 +
      labs(x = "date") +
      theme(axis.text.y = element_blank())) +
    (violin_2 +
      labs(y = "anti-S IgG") +
      scale_y_continuous(trans = "log10") +
      annotation_logticks(sides = "l"))
) +
  plot_layout(design = lay2) &
  plot_annotation(tag_levels = "A") &
  theme_bw(base_size = 8) &
  theme(legend.position = "none") &
  theme(plot.tag = element_text(size = 16, face = "bold"))
```

Note that for participant 3, the sample \~50d before dose 2 is now excluded.

## Summary

Chronogram is now assembled and the relevant data to answer a (simple) biological question (does dose 2 boost anti-S IgG?) retrieved, plotted and readied for a statistical test.

Please see the other vignettes for a deeper demonstration of chronogram functions.

## SessionInfo

```{r}
sessionInfo()
```
