---
title: "The Whole Game"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Whole Game}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```


This vignette is for the impatient and showcases the features of the chronogram package.

## Load libraries

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(chronogram)
```


## Construct chronogram

```{r}
## Fictional example data ##
data(smallstudy)


ids <- smallstudy$small_study_ids
start <- "01012020"
end <- "10102021"
meta <- smallstudy$small_study_metadata
ab <- smallstudy$small_study_Ab

small_study <- chronogram_skeleton(
  ids = ids,
  start_date = start,
  end_date = end,
  ## change this to your ID column name ##
  col_ids = elig_study_id,
  ## change this to your date column name ##
  col_calendar_date = calendar_date
)

small_study <- chronogram(small_study, meta)

small_study <- cg_add_experiment(small_study, ab)
```

## Plot trajectories

Plot antibody responses over time for each individual, with a swimmer plot to summarise each individuals course.

```{r}
traj <- cg_plot(small_study,
  y_values = serum_Ab_S
)

swimmer <- cg_plot_meta(small_study,
  visit = serum_Ab_S
)

traj
swimmer
```

### Customise plots

Using ggplot2 or patchwork adjustments.

```{r}
lay <- "
a
a
a
b
"
(
  (traj + labs(y = "anti-S IgG") + scale_y_log10()) /
    (swimmer + theme(axis.text.y = element_blank()))
) +
  plot_layout(design = lay) &
  plot_annotation(tag_levels = "A") &
  theme_bw(base_size = 8) &
  theme(legend.position = "none") &
  theme(plot.tag = element_text(size = 16, face = "bold")) &
  labs(x = "date")
```

## Subset data

```{r}
## upto 50d after dose 2 ##
cg_after_dose_2 <- cg_window_by_metadata(small_study,
  windowing_date_col = date_dose_2,
  preceding_days = 0,
  following_days = 50
)

## upto 21d before dose 2 ##
cg_before_dose_2 <- cg_window_by_metadata(small_study,
  windowing_date_col = date_dose_2,
  preceding_days = 21,
  following_days = 0
)


cg_after_dose_2 <- cg_after_dose_2 %>% mutate(cohort = "post D2")

cg_before_dose_2 <- cg_before_dose_2 %>% mutate(cohort = "pre D2")

cg_dose_2 <- bind_rows(cg_after_dose_2, cg_before_dose_2)

cg_dose_2 <- cg_dose_2 %>% mutate(
  cohort =
    factor(cohort,
      levels = c(
        "pre D2",
        "post D2"
      )
    )
)
```

### Re-plot the subsetted data

```{r}
traj_2 <- cg_dose_2 %>%
  cg_plot(.,
    y_values = serum_Ab_S
  )

pd <- position_dodge(0.4)

violin_2 <- cg_dose_2 %>%
  ## geom_line fails if lots of NA rows provided ##
  filter(!is.na(serum_Ab_S)) %>%
  ggplot(aes(
    x = cohort, y = serum_Ab_S,
    group = elig_study_id
  )) +
  geom_point(position = pd) +
  geom_line(position = pd)

swimmer_2 <- cg_plot_meta(cg_dose_2,
  visit = serum_Ab_S
)


lay2 <- "
aac
aac
aa#
bb#
"

(
  (traj_2 +
    labs(y = "anti-S IgG", x = "date") +
    scale_y_log10()) +
    (swimmer_2 +
      labs(x = "date") +
      theme(axis.text.y = element_blank())) +
    (violin_2 +
      labs(y = "anti-S IgG") +
      scale_y_continuous(trans = "log10") +
      annotation_logticks(sides = "l"))
) +
  plot_layout(design = lay2) &
  plot_annotation(tag_levels = "A") &
  theme_bw(base_size = 8) &
  theme(legend.position = "none") &
  theme(plot.tag = element_text(size = 16, face = "bold"))
```

Note that for participant 3, the sample ~50d before dose 2 is now excluded.

## Summary

Chronogram is now assembled and the relevant data to answer a (simple) biological question (does dose 2 boost anti-S IgG?) retrieved, plotted and readied for a statistical test.

Please see the other vignettes for a deeper demonstration of chronogram functions.

## SessionInfo

```{r}
sessionInfo()
```

