---
title: "Chronogram statistics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chronogram statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

In this vignette, we explore how to prepare chronogram data for statistical testing. We take a provided assembled chronogram and perform filtering, windowing and selecting prior to an illustrative plot and a statistical test. This vignette does not cover when each statistical test is appropriate, nor the R functions (or arguments) for each particular statistical test.

### Libraries

```{r setup}
library(dplyr)
library(rstatix) # pipe-friendly stats functions
library(broom) # tidy handling of lm() output
library(ggrepel) # repelled labels in ggplot2
library(survival) # time-to-event analysis
library(survminer) # ggplot2 from time-to-event
library(chronogram)
```

## Wilcoxon test

-   *Hypothesis*

    For individuals without prior exposure, a long interval (8-12w) between doses 1 and 2 of an mRNA vaccine will have higher antibody levels 3-5w after dose 2, than individuals following a short interval (3-4w).

-   *response variable*

    MSD SARS-CoV-2 Spike antibody (`CoV_2S_MSD`).

-   *categorical predictor*

    `LONG` or `SHORT` interval (`Vaccine_interval` in metadata).

-   *data*

    PITCH, introduced and assembled in the [SQL vignette](articles/SQL_assembly.html).

```{r}
data("pitch_chronogram")

## filter, select, window ##
windowed_data <- pitch_chronogram %>%
  filter(Previous_infection == "N") %>%
  cg_window_by_metadata(dose_2_date, 
                        preceding_days = -21, 
                        following_days = 35) %>%
  cg_select_visit(visit_col = Cov_2S_BAU,
                  visit = "latest")

```

```{r}
## plot ##
windowed_data %>%
  ggplot(aes(x = Vaccine_interval, y = Cov_2S_MSD)) +
  geom_point(shape = 20, alpha = 0.4,
             position = position_jitter(0.3)) +
  stat_summary(
    geom = "point",
    fun = "median",
    shape = 5,
    size = 2,
    stroke = 1.5,
    col = "dodgerblue") +
  theme_bw()

## stats ##
windowed_data %>%
  wilcox_test(Cov_2S_MSD ~ Vaccine_interval) ## rstatix::wilcox_test

```

------------------------------------------------------------------------

## Linear regression

-   *Hypothesis*

    SARS-CoV-2 Spike antibody is dependent on vaccine interval, previous infection, and not on study centre.

-   *response variable*

    MSD SARS-CoV-2 Spike (`CoV_2S_MSD`).

-   *predictors*

    -   `LONG` or `SHORT` interval (`Vaccine_interval` in metadata).

    -   study centre (`Centre_code` in metadata).

    -   Previous infection (`Y`/`N` in `Previous_infection` in metadata).

-   *data*

    PITCH, introduced and assembled in the [SQL vignette](articles/SQL_assembly.html).

```{r}
data("pitch_chronogram")

## filter, select, window ##
windowed_data <- pitch_chronogram %>%
  cg_window_by_metadata(dose_2_date, 
                        preceding_days = -21, 
                        following_days = 35) %>%
  cg_select_visit(visit_col = Cov_2S_BAU,
                  visit = "latest")

```

```{r}
## plot ##
windowed_data %>%
  ggplot(aes(x = Centre_code, y = Cov_2S_MSD)) +
  geom_point(shape = 20, alpha = 0.4,
             position = position_jitter(0.3)) +
  stat_summary(
    geom = "point",
    fun = "median",
    shape = 5,
    size = 2,
    stroke = 1.5,
    col = "dodgerblue") +
  facet_grid(. ~Vaccine_interval + Previous_infection ) +
  theme_bw()

## stats ##
lm_res <- windowed_data %>%
  lm(Cov_2S_MSD ~ 
       Vaccine_interval + 
       Previous_infection + 
       Centre_code, data = .)

## results from lm ##
summary(lm_res)

## easier to read, and plot from broom::tidy ##
tidy(lm_res)

tidy(lm_res) %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x=estimate, y = -log10(p.value),
         label = term)) +
  geom_point() +
  geom_text_repel() +
  geom_hline(yintercept = -log10(0.05), 
             linetype = 3)

```

------------------------------------------------------------------------

### Linear regression with interaction terms

Setup as for `Linear regression` example above.

Since nearly all of the `SHORT` interval data were from a single centre, we add an interaction term between centre and interval.

```{r}
## stats - lm with interaction terms ##
## note the *
lm_res_interactions <- windowed_data %>%
  lm(Cov_2S_MSD ~ 
       Previous_infection +
       Vaccine_interval * 
       Centre_code, data = .)

## results from lm ##
summary(lm_res_interactions )

## easier to read, and plot from broom::tidy ##
tidy(lm_res_interactions )

tidy(lm_res_interactions ) %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x=estimate, y = -log10(p.value),
         label = term)) +
  geom_point() +
  geom_text_repel() +
  geom_hline(yintercept = -log10(0.05), 
             linetype = 3)
```

The model `Cov_2S_MSD ~ Previous_infection + Vaccine_interval * Centre_code`, with an interaction term between interval and centre performs reasonably, with a neglible effect assigned to centre alone, and larger effects assigned to vaccine interval and previous infection.

By ANOVA, `Cov_2S_MSD ~ Previous_infection + Vaccine_interval * Centre_code` performs better than `Cov_2S_MSD ~ Previous_infection + Vaccine_interval + Centre_code`:

```{r}
anova(lm_res, 
      lm_res_interactions)
```

Either model describes \~10% of the total variance in Spike antibodies:

```{r}

bind_rows(
  # this is broom::glance() : #
  glance(lm_res),
  glance(lm_res_interactions)
)

```

------------------------------------------------------------------------

## Time to event analysis

-   *Hypothesis*

    Within centre 1, dose 2 ("event") occurs earlier in individuals without prior infection.

-   *Start*

    date of dose 1

-   *Event*

    date of dose 2

-   *Censoring*

    none (all of PITCH have both doses)

-   *Filtering*

    to centre 0, as that centre has an approximately equal split of `long`:`short` interval.

> **NOTE**
>
> Changes have been made to the de-identified public dataset (DOI: 10.17632/fyp26zjgmj.1).
>
> i.  the public data reports dates as MM/YYYY. To build a chronogram we have assigned randomly from 1-28 for DD in DD/MM/YYYY for dose 1, and used the available intervals in days to place the remaining data in date time. For illustration purposes, we need plausible DD/MM/YYYY dates - they are not real.
>
> ii. not all of the public dataset has been used in this example (some assays not included).

For meaningful time to event analysis, **real** rather than **plausible** dates are needed. This vignette shows example code, but the results themselves are insecure.

```{r}
data("pitch_chronogram")

## filter, de-duplicate ##
t_to_event_data <- pitch_chronogram %>%
  filter(Centre_code == 0) %>%
  group_by(ID) %>%
  slice_head() %>%
  ungroup() %>%
  ## all receive dose 2 ##
  mutate(outcome = 1) %>%
  mutate(dose_2_crude = 
            as.numeric(
           dose_2_date - (dose_1_date)))

fit<- survfit(Surv(time = dose_2_crude,
                   # time2 = dose_2_d,
                   outcome) ~ Previous_infection, data = t_to_event_data)

ggsurvplot(fit,
                      conf.int = TRUE,
                      data = t_to_event_data,
                      fun = "event",
                      risk.table = TRUE,
                      pval = TRUE,
                      pval.coord = c(0.1, 0.9),
                      pval.method = TRUE,
                      pval.method.coord = c(0.1, 1))



```

### Left-censored time to event analysis

Here we use left censoring, such that a given individual is only considered at-risk for the event (dose 2) once they have had their first dose.

In the real world, left censoring is useful if the risk of exposure might change over calendar time. In this example dataset, it highlights the use of *plausible* (DD randomly assigned to MM-YYYY), rather than *real* dates.

```{r}
data("pitch_chronogram")

## filter, de-duplicate ##
t2_to_event_data <- pitch_chronogram %>%
  filter(Centre_code == 0) %>%
  group_by(ID) %>%
  slice_head() %>%
  ungroup() %>%
  ## all receive dose 2 ##
  mutate(outcome = 1) %>%
  ## use the earliest date of dose 1 to anchor other times ##
  mutate(dose_1_d = 
            as.numeric(
           dose_1_date - min(dose_1_date))) %>%
  mutate(dose_2_d = 
            as.numeric(
           dose_2_date - min(dose_1_date)))


fit2 <- survfit(Surv(time = dose_1_d,
                    time2 = dose_2_d,
                   outcome) ~ Previous_infection, data = t2_to_event_data)



ggsurvplot(
  fit2,
  conf.int = TRUE,
  data = t2_to_event_data,
  fun = "event",
  risk.table = TRUE,
  ## calcuate Cox PH P value ##
  pval = 
    round(
    summary(
    coxph(Surv(time = dose_1_d,
               time2 = dose_2_d,
               outcome) ~ Previous_infection, 
          data = t2_to_event_data)
    )$coefficients[,"Pr(>|z|)"],
    digits = 3),
  pval.coord = c(0.1, 0.9)
  # pval.method.coord = c(0.1, 1)
)


```

Note:

1.  now the x axis runs to over 5 months as it covers the \~2m period between the earliest and latest first dose.

2.  the numbers at risk can increase between time points as more individuals become eligible (here, by receiving dose 1).

The difference between strata is now much less apparent. This is likely due to our *plausible* rather than *real* dates in the example dataset.

------------------------------------------------------------------------

## Summary

This vignette shows code to map data from a chronogram to a variety of different statistical tests.

------------------------------------------------------------------------

## SessionInfo

```{r}
sessionInfo()
```
