---
title: "Symptoms_diary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Symptoms_diary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(chronogram)
library(dplyr)
library(ggplot2)
```

We re-use the `smallstudy` dataset with two modifications:

- addition of participant #4, who has only dose 1

- addition of symptoms diary information, including that participant #4 had a serious reaction to vaccination and therefore opted not to have further vaccinations.

> This is a plausible, but fictitious dataset.

```{r}
data(smallstudy)
metadata <- smallstudy$small_study_metadata
ab <- smallstudy$small_study_Ab


## Add a 4th participant's metadata ##
## [in real use, a bigger metadata tibble should be made, not two tibbles combined via bind_rows()]
metadata  <- bind_rows(
  metadata,
  tibble::tribble(
      ~elig_study_id,   ~age, ~sex,   ~dose_1,   ~date_dose_1, ~dose_2,   ~date_dose_2,
      4,             25,    "M",   "AZD1222", lubridate::dmy('07012021'), NA, NA
)
)

metadata <- metadata %>%
  mutate(sex = factor(sex)) %>%
  mutate(dose_1 = factor(dose_1))

metadata

##--------------------------------------------------------------------
## Add symptoms diary info

symptoms_to_add <- tibble::tribble(
  ~elig_study_id, ~calendar_date, ~symptom_fever, ~symptom_headache, ~symptom_joint_pain, ~symptom_rash,
  1,              "06012021",     "mild",          "moderate",           NA,                  NA,
  1,              "07012021",     "mild",              NA ,              NA,                  NA,
  1,              "08012021",       NA,                NA ,            "severe",              NA,
  4,              "10012021",       NA,                 NA,               NA,          "required_hospitalisation")

symptoms_to_add <- symptoms_to_add %>%
  mutate(calendar_date = lubridate::dmy(calendar_date))

```

Assemble this new chronogram

```{r}
start <- "01012020"
end <- "10102021"

cg <- cg_assemble(
 start_date = start,
 end_date = end,
 calendar_date_col = calendar_date,
 metadata = metadata,
 metadata_ids_col = elig_study_id,
 experiment_data_list = list(ab, symptoms_to_add)
)


```

First annotate the vaccines

```{r}
cg <- cg_annotate_vaccines_count(
  cg,
  ## the prefix to the dose columns: ##
  dose = dose,
  ## the output column name: ##
  dose_counter = dose_number,
  ## the prefix to the date columns: ##
  vaccine_date_stem = date_dose,
  ## use 14d to 'star' after a dose ##
  intermediate_days = 14
)

## plot over time ##
cg %>%
  ## refactor for gradient fill ##
  mutate(dose_number = factor(dose_number,
    levels = c(
      "0",
      "1star",
      "1",
      "2star",
      "2"  ))) %>%
  ggplot(
    aes(
      x = calendar_date,
      y = elig_study_id,
      fill = dose_number
    )
  ) +
  geom_tile(height = 0.5) +
  scale_fill_grey(end = 0.2, start = 0.8) +
  theme_bw()

```

Now we can plot how these symptoms to see how they map in time to vaccinations.

```{r}

cg %>%
  ## refactor for gradient fill ##
  mutate(dose_number = factor(dose_number,
    levels = c(
      "0",
      "1star",
      "1",
      "2star",
      "2" ))) %>%
  ggplot(
    aes(
      x = calendar_date,
      y = elig_study_id,
      fill = dose_number
    )
  ) +
  geom_tile(height = 0.3) +
  scale_fill_grey(end = 0.2, start = 0.8) +
   geom_point(data = . %>% 
               ## filter to symptoms that are present ##
               filter(!is.na(symptom_rash)) %>% 
               mutate(symptom_rash = paste("rash", symptom_rash)),
             aes(col = symptom_rash),
             shape = "I",
             size = 4,
             position = position_nudge(y=0.2)) +
   geom_point(data = . %>% 
               
               filter(!is.na(symptom_headache)) %>% 
               
               mutate(symptom_headache = paste("headache", symptom_headache)),
             aes(col = symptom_headache),
             shape = "I",
             size = 4,
             position = position_nudge(y=0.2)) +
  geom_point(data = . %>% 
               
               filter(!is.na(symptom_fever)) %>% 
               
               mutate(symptom_fever = paste("fever", symptom_fever)),
             aes(col = symptom_fever),
             shape = "I",
             size = 4,
             position = position_nudge(y=0.3)) +
   geom_point(data = . %>% 
               
               filter(!is.na(symptom_joint_pain)) %>% 
               
               mutate(symptom_joint_pain = paste("joint pain", symptom_joint_pain)),
             aes(col = symptom_joint_pain),
             shape = "I",
             size = 4,
             position = position_nudge(y=0.4)) +
  ## swap the fill scale, and stop colour being included in this guide ##
  scale_fill_grey(
    end = 0.2, start = 0.8,
    name = "dose_number",
                  na.translate = FALSE,
                  na.value = NA, 
                  
                  guide = guide_legend(override.aes = list(color = NA))) +
  ## swap the colour scale ##
  scale_color_brewer(type = "qual",palette = 2,
                     name = "symptoms") + 
  theme_bw() + 
  theme(legend.position = "right",
        legend.direction = "vertical")
  
```

In our test dataset:

- participant 1 reported a post-first-dose fever, headache, and joint pain a few days later.

- participants 2 and 3 reported no symptoms after either dose.

- participant 4 reported a significant symptom after dose 1 and did not receive further doses.


## SessionInfo

```{r}
sessionInfo()
```
